% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  letterpaper,
]{svmono}

\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
    \setmainfont[]{Palatino}
    \setmonofont[]{Inconsolata}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{5}
% Make \paragraph and \subparagraph free-standing
\makeatletter
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}{
    \@ifstar
      \xxxParagraphStar
      \xxxParagraphNoStar
  }
  \newcommand{\xxxParagraphStar}[1]{\oldparagraph*{#1}\mbox{}}
  \newcommand{\xxxParagraphNoStar}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}{
    \@ifstar
      \xxxSubParagraphStar
      \xxxSubParagraphNoStar
  }
  \newcommand{\xxxSubParagraphStar}[1]{\oldsubparagraph*{#1}\mbox{}}
  \newcommand{\xxxSubParagraphNoStar}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
\makeatother


\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\newsavebox\pandoc@box
\newcommand*\pandocbounded[1]{% scales image to fit in text height/width
  \sbox\pandoc@box{#1}%
  \Gscale@div\@tempa{\textheight}{\dimexpr\ht\pandoc@box+\dp\pandoc@box\relax}%
  \Gscale@div\@tempb{\linewidth}{\wd\pandoc@box}%
  \ifdim\@tempb\p@<\@tempa\p@\let\@tempa\@tempb\fi% select the smaller of both
  \ifdim\@tempa\p@<\p@\scalebox{\@tempa}{\usebox\pandoc@box}%
  \else\usebox{\pandoc@box}%
  \fi%
}
% Set default figure placement to htbp
\def\fps@figure{htbp}
\makeatother
% definitions for citeproc citations
\NewDocumentCommand\citeproctext{}{}
\NewDocumentCommand\citeproc{mm}{%
  \begingroup\def\citeproctext{#2}\cite{#1}\endgroup}
\makeatletter
 % allow citations to break across lines
 \let\@cite@ofmt\@firstofone
 % avoid brackets around text for \cite:
 \def\@biblabel#1{}
 \def\@cite#1#2{{#1\if@tempswa , #2\fi}}
\makeatother
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[2] % #1 hanging-indent, #2 entry-spacing
 {\begin{list}{}{%
  \setlength{\itemindent}{0pt}
  \setlength{\leftmargin}{0pt}
  \setlength{\parsep}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
   \setlength{\leftmargin}{\cslhangindent}
   \setlength{\itemindent}{-1\cslhangindent}
  \fi
  % set entry spacing
  \setlength{\itemsep}{#2\baselineskip}}}
 {\end{list}}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{\hfill\break\parbox[t]{\linewidth}{\strut\ignorespaces#1\strut}}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

% template.tex: Springer-compatible LaTeX header for Quarto

\usepackage{fontspec}

% Essential Springer-required fonts and packages
% \usepackage{mathptmx} % Times font (Springer default)
% \usepackage{helvet}   % Helvetica font (for sans-serif text)
% \usepackage{courier}  % Courier font (monospaced text)
% \usepackage{type1cm}  % Ensures scalable fonts (Type 1)

% Additional Springer-specific settings:
\usepackage{makeidx}  % Required for indexing
\makeindex            % Enable index generation

% For better tables
\usepackage{booktabs}

% Figures and graphics settings
\usepackage{graphicx}

% Math packages (often used in Springer books)
\usepackage{amsmath}
\usepackage{amssymb}

% Springer-specific gray boxes (optional):
\usepackage{tcolorbox}
\tcbuselibrary{listingsutf8}
\tcbset{
  boxrule=0pt,
  colback=gray!10,
  colframe=gray!40,
  sharp corners
}

\usepackage{listings}
\lstset{
  breaklines=true,
  breakatwhitespace=false,
  postbreak=\mbox{\textcolor{gray}{$\hookrightarrow$}\space},
  basicstyle=\monofont,
  columns=fullflexible,
  keepspaces=true,
  frame=single,
  xleftmargin=1em,
  tabsize=2
}


\usepackage{fvextra}
\DefineVerbatimEnvironment{Highlighting}{Verbatim}{
  breaklines=true,
  breakanywhere=true,
  fontsize=\small,
  commandchars=\\\{\}
}

\numberwithin{example}{subsection}


% Custom commands (optional):
\newcommand{\R}{\textsf{R}}

% Epigraph environment
\newenvironment{chapterquote}
  {\begin{quote}\itshape}
  {\end{quote}\vspace{2em}}

%%%%%%%%%%%%%%
% --- Make sure the TOC does NOT add itself as an entry ---
\makeatletter
\let\origtableofcontents\tableofcontents
\renewcommand{\tableofcontents}{%
  \begingroup
  \let\addcontentsline\@gobblethree
  \let\addtocontents\@gobbletwo
  \origtableofcontents
  \endgroup
}
\makeatother



\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\makeatletter
\@ifpackageloaded{bookmark}{}{\usepackage{bookmark}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother

\usepackage{bookmark}

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Data Science Foundations and Machine Learning with R: From Data to Decisions},
  pdfauthor={Reza Mohammadi},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}


\title{{Data Science Foundations and Machine Learning with R: From Data
to Decisions}}
\author{\href{https://www.uva.nl/profile/a.mohammadi}{{Reza Mohammadi}}}
\date{5 May 2026}

\begin{document}
\maketitle

\renewcommand*\contentsname{Table of contents}
{
\hypersetup{linkcolor=}
\setcounter{tocdepth}{2}
\tableofcontents
}

\bookmarksetup{startatroot}

\chapter*{}\label{section}
\addcontentsline{toc}{chapter}{}

\markboth{}{}

\clearpage
\thispagestyle{empty}

\vspace*{0.33\textheight}

\begin{center}
\begin{minipage}{0.65\textwidth}
\centering
\emph{This book is dedicated to those in Iran who lost their lives in 2026 in the hope of freedom.}
\end{minipage}
\end{center}

\clearpage

\bookmarksetup{startatroot}

\chapter*{Preface}\label{preface}
\addcontentsline{toc}{chapter}{Preface}

\markboth{Preface}{Preface}

Many students encounter data science through fragments: a statistical
formula in one course, an R command in another, and a machine learning
algorithm presented as a ready-made tool. They may learn how to run code
without fully understanding the method, or they may study concepts
without seeing how those concepts guide real data analysis. This
separation between statistical reasoning, programming, and machine
learning can make the field appear more technical and less coherent than
it needs to be. An introductory textbook should therefore do more than
present methods one by one. It should help readers understand how
analytical questions, data preparation, statistical thinking, model
building, evaluation, and interpretation fit together within a
reproducible workflow.

\emph{Data Science Foundations and Machine Learning with R: From Data to
Decisions} was written to support this integrated view of data science
education. The book is designed for readers with no prior experience in
analytics, programming, or formal statistics, including students,
professionals, and researchers seeking a practical introduction to data
science and machine learning. Drawing on classroom experience, it
combines foundational statistical concepts, modern machine learning
methods, reproducible R workflows, and realistic applications. Concepts
are introduced progressively and reinforced through examples and applied
tasks, with the aim of developing both conceptual understanding and
practical fluency. The central goal is to make data science accessible
to beginners while preserving the rigor needed to apply methods
critically, reproducibly, and responsibly.

\section*{Motivation and Scope}\label{motivation-and-scope}

\markright{Motivation and Scope}

The book addresses this educational challenge by treating data science
as an integrated process rather than as a sequence of isolated topics.
Statistical reasoning, programming, visualization, data preparation, and
machine learning are introduced as complementary parts of applied
analysis. This organization allows readers to see not only how
individual methods work, but also how they contribute to complete
data-driven investigations.

The book differs from purely theoretical introductions by placing
methods in practical analytical contexts, and it differs from software
manuals by treating R as a means for statistical reasoning rather than
as an end in itself. Each chapter introduces concepts gradually and
illustrates them through reproducible R examples and applied analyses.
This structure is intended to help readers move from following examples
to carrying out their own analyses with greater independence.

R was chosen because of its central role in statistical computing, data
visualization, reproducible research, and applied data science
education. Its open-source ecosystem allows readers to work with
professional tools while developing transparent and reproducible
workflows. Throughout the book, R is used not only to implement methods,
but also to support interpretation, model evaluation, and communication
of results.

The scope of the book is introductory but broad. It covers the
foundations of data science, data preparation, exploratory data
analysis, statistical inference, supervised learning, model evaluation,
regression models, classification methods, tree-based models, neural
networks, and clustering. The aim is not to provide exhaustive treatment
of every method, but to build a principled foundation that prepares
readers to apply data science and machine learning methods critically in
academic, professional, and research settings.

\section*{Audience and Prerequisites}\label{audience-and-prerequisites}

\markright{Audience and Prerequisites}

The primary audience for this book is undergraduate and early graduate
students taking an introductory course in data science, machine
learning, business analytics, applied statistics, econometrics, or
quantitative methods. The book is particularly suitable for programs in
which students need to develop practical data analysis skills while also
understanding the statistical and computational ideas behind the methods
they use.

The book is designed for readers who are new to data science,
programming, and machine learning. No prior experience with R is
assumed, and the necessary programming concepts are introduced from the
beginning. A formal background in statistics is also not required,
although readers are expected to engage with statistical ideas as they
are developed throughout the book, especially in the chapters on
exploratory data analysis, statistical inference, regression, and model
evaluation.

The material is also suitable for professionals, researchers, and
self-study readers who wish to build a practical foundation in data
science and machine learning using R. Readers from fields such as
business, economics, the social sciences, communication science,
psychology, health sciences, and STEM disciplines can use the book to
develop reproducible analytical workflows and to apply data-driven
methods to realistic problems.

Successful use of the book requires active engagement rather than
passive reading. Readers are encouraged to run the R code, inspect
intermediate results, modify examples, complete the in-text
\emph{Practice} tasks, and work through the end-of-chapter exercises.
Basic computer literacy and access to a computer capable of running R
and RStudio are sufficient. Installation instructions and the required
software setup are introduced in Chapter \textbf{?@sec-ch1-intro-R}.

\section*{Learning Goals}\label{learning-goals}

\markright{Learning Goals}

This book is designed to help readers develop both conceptual
understanding and practical skill in data science and machine learning
with R. The goals below reflect the progression of the book, from
formulating analytical questions to preparing data, building models,
evaluating results, and communicating findings.

By the end of the book, readers will be able to:

\begin{itemize}
\item
  formulate data science questions and relate them to the main stages of
  the Data Science Workflow, from problem understanding to evaluation
  and interpretation;
\item
  use R to import, inspect, transform, visualize, and analyze data
  within a reproducible workflow;
\item
  prepare and explore datasets by identifying variable types, addressing
  missing values and outliers, transforming variables, and using
  graphical and numerical summaries;
\item
  apply core ideas from statistical inference and model evaluation to
  assess uncertainty, compare results, and judge model performance in
  context;
\item
  build, tune, and evaluate supervised and unsupervised learning models,
  including regression models, classification methods, tree-based
  models, neural networks, and clustering methods;
\item
  interpret analytical results critically and communicate findings in a
  way that supports evidence-based decisions.
\end{itemize}

Together, these goals emphasize not only technical implementation, but
also interpretation, evaluation, and responsible use of data science
methods. The aim is to help readers understand when, why, and how data
science tools should be applied.

\section*{Pedagogical Approach}\label{pedagogical-approach}

\markright{Pedagogical Approach}

The book is designed to support active learning in self-study, classroom
instruction, and professional training. Readers may work through the
chapters sequentially or consult selected chapters for specific topics,
but the material is most effective when studied interactively. Concepts
are introduced through explanation and examples, then reinforced through
annotated R code, in-text \emph{Practice} tasks, case studies, and
end-of-chapter exercises.

Readers are encouraged to run the code, inspect intermediate results,
modify examples, and compare alternative choices such as different
variables, parameter settings, or models. The case studies show how
individual methods contribute to complete analyses, while the exercises
provide further opportunities to consolidate understanding and develop
fluency in R. This approach emphasizes reproducibility throughout:
readers learn not only how to obtain results, but also how to document,
check, and interpret the analytical steps that produced them.

\section*{Structure of This Book}\label{structure-of-this-book}

\markright{Structure of This Book}

The book is organized around a Data Science Workflow, developed in
Chapter \textbf{?@sec-ch2-intro-data-science} and revisited throughout
the text. This workflow provides a practical structure for moving from
problem formulation and data preparation to modeling, evaluation, and
interpretation. Rather than presenting data science as a collection of
separate techniques, the chapters show how statistical reasoning,
computation, and machine learning contribute to a coherent analytical
process.

Chapters \textbf{?@sec-ch1-intro-R} and
\textbf{?@sec-ch2-intro-data-science} establish the foundations. Chapter
\textbf{?@sec-ch1-intro-R} introduces R, RStudio, basic programming
concepts, data structures, data import, and reproducible reporting.
Chapter \textbf{?@sec-ch2-intro-data-science} introduces the main ideas
of data science and machine learning, including the Data Science
Workflow that guides the rest of the book.

Chapters \textbf{?@sec-ch3-data-preparation} to
\textbf{?@sec-ch6-data-setup} focus on preparing readers for modeling.
Chapter \textbf{?@sec-ch3-data-preparation} examines data preparation,
including variable types, missing values, outliers, and transformations.
Chapter \textbf{?@sec-ch4-EDA} introduces exploratory data analysis
through numerical summaries and visualization. Chapter
\textbf{?@sec-ch5-statistics} develops core ideas from statistical
inference and hypothesis testing, which support uncertainty assessment
and critical interpretation. Chapter \textbf{?@sec-ch6-data-setup} then
turns to data setup for modeling, including partitioning, validation,
preprocessing, leakage prevention, and class imbalance.

Chapters \textbf{?@sec-ch7-classification-knn} to
\textbf{?@sec-ch11-generalized-regression} introduce supervised learning
and regression-based modeling. These chapters cover classification with
k-nearest neighbors, model evaluation, Naive Bayes classification,
regression analysis for continuous outcomes, and generalized regression
models for binary and count outcomes. Together, they show how predictive
models are built, assessed, compared, and interpreted in applied
settings.

Chapters \textbf{?@sec-ch12-tree-models} to
\textbf{?@sec-ch14-clustering} extend the modeling toolkit. Chapter
\textbf{?@sec-ch12-tree-models} introduces decision trees and random
forests, Chapter \textbf{?@sec-ch13-neural-networks} presents neural
networks for supervised learning, and Chapter
\textbf{?@sec-ch14-clustering} introduces clustering as an unsupervised
learning approach. These chapters broaden the methodological scope while
continuing to emphasize practical implementation, evaluation, and
interpretation.

Each chapter combines conceptual explanation with reproducible R
examples, applied tasks, a realistic case study, and end-of-chapter
exercises. This structure allows readers to encounter each method first
as an idea, then as an implementation, and finally as part of a complete
analysis. The emphasis throughout is on developing reproducible
workflows and using data science methods critically in context.

\section*{For Instructors}\label{for-instructors}

\markright{For Instructors}

This book can be used as a primary or supplementary text in introductory
courses on data science, machine learning, business analytics, applied
statistics, econometrics, and quantitative methods. Its structure
supports use in undergraduate courses, early graduate courses,
professional training programs, and self-contained practical modules.

The chapters are designed to support flexible teaching. Instructors may
follow the full sequence of the book or select chapters according to the
aims of a specific course. The book includes an extensive collection of
exercises, ranging from conceptual questions to applied coding tasks and
more advanced problems. Each chapter also includes a case study that can
be used for classroom discussion, practical sessions, assignments, or
project-based learning.

Additional teaching materials, including lecture slides, practical data
science projects, and assessment resources, are made available to
support course preparation and delivery. Further information about
instructor resources and updates is available at the book's companion
website: \url{https://datasciencebook.ai}.

\section*{Software, Data, and Online
Resources}\label{software-data-and-online-resources}

\markright{Software, Data, and Online Resources}

The book uses real-world datasets to support hands-on learning and
reproducible analysis. These datasets are used throughout the chapters
to illustrate concepts, demonstrate R workflows, and develop case
studies in areas such as customer analytics, finance, marketing,
healthcare, housing, and classification under class imbalance. Most
datasets used in the book are provided through the \textbf{liver}
package, which allows readers to reproduce examples, complete exercises,
and extend the analyses without needing to locate external data sources
before beginning the work.

The \textbf{liver} package contains datasets used in examples,
exercises, and case studies, including datasets such as \texttt{churn},
\texttt{bank}, \texttt{adult}, \texttt{loan},
\texttt{purchase\_intention}, \texttt{doctor\_visits},
\texttt{wholesale\_customers}, \texttt{creditcard\_fraud},
\texttt{red\_wines}, and \texttt{white\_wines}. The package also
includes supporting functions used throughout the book for common data
science tasks. Datasets can be loaded directly in R using the
\texttt{data()} function, and their documentation provides descriptions
of the variables and references to the original sources where
applicable.

Additional resources are available through the book's companion website,
\url{https://datasciencebook.ai}. The website provides updates,
supplementary materials, and information about resources for instructors
and readers. Documentation for the \textbf{liver} package is available
through CRAN at
\url{https://cran.r-project.org/web/packages/liver/index.html},
including dataset descriptions and reference material.

For readers interested in a Python-based treatment of the same learning
objectives, a companion volume, \emph{Data Science Foundations and
Machine Learning with Python: From Data to Decisions}, is available from
the same publisher. Further information about both books and their
supporting resources is provided on the companion website.

\section*{Acknowledgments}\label{acknowledgments}

\markright{Acknowledgments}

Writing this book has been a demanding and rewarding process, and I am
grateful to the many people who supported its development. First, I
thank my wife, Pariya, for her patience, encouragement, and constant
support throughout this project. I am also deeply grateful to my family,
especially my mother and older brother, for their belief in me.

I am especially thankful to Eva Hiripi at Springer for her support and
encouragement from the early stages of this book. I also thank Dr.~Kevin
Burke for his valuable input on the structure of the book, and
Dr.~Jeroen van Raak and Dr.~Julien Rossi for their collaboration on the
Python edition.

I am grateful to my colleagues in the Business Analytics Section at the
University of Amsterdam for their feedback, encouragement, and academic
support during the writing process. In particular, I thank Prof.~Ilker
Birbil, Prof.~Dick den Hertog, Prof.~Marc Salomon, Dr.~Marit
Schoonhoven, Dr.~Stevan Rudinac, Dr.~Rob Goedhart, Prof.~Jeroen de Mast,
Prof.~Joaquim Gromicho, Prof.~Peter Kroos, Dr.~Chintan Amrit, Dr.~Inez
Zwetsloot, Dr.~Alex Kuiper, Dr.~Bart Lameijer, Dr.~Jannis Kurtz,
Dr.~Guido van Capelleveen, and Dr.~Yeqiu Zheng. I also thank my PhD
students, Lucas Vogels and Elias Dubbeldam, for their research insights
and continued collaboration.

I further acknowledge my former colleagues and co-authors,
Dr.~Khodakaram Salimifard, Sara Saadatmand, and Dr.~Florian
Böing-Messing, for their continued academic partnership. Finally, I am
grateful to the students of the courses \emph{Data Wrangling} and
\emph{Data Analytics: Machine Learning} at the University of Amsterdam.
Their questions, feedback, and engagement helped refine the material in
meaningful ways. I am particularly thankful to John Gatev for his
thoughtful and constructive comments.

\begin{center}
\textit{All models are wrong, but some are useful.}
\end{center}
\begin{flushright}
— George Box
\end{flushright}

Reza Mohammadi\\
Amsterdam, Netherlands\\
May 2026

\bookmarksetup{startatroot}

\chapter*{References}\label{references}
\addcontentsline{toc}{chapter}{References}

\markboth{References}{References}

\phantomsection\label{refs}
\begin{CSLReferences}{0}{1}
\end{CSLReferences}




\end{document}
